---
title: 'MB5370 Module One: Introduction to Programming'
author: "Caitlin Buckle"
date: "2026-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#   Workshop One: Introduction (Wed 20th May, 9-1pm)
### NOTES TO MYSELF 

Set-up an R Project and store any/all data/Rmd files within the Project 

Use getwd() to see where the particular file is stored on the computer (shouldn't need to use this if correctly using a R Project)

When reopening an R project, do not directly open the Rmd file (you will be able to edit the file but it will not save as part of the project). Check the right-hand top corner as to which project you are working in : ) 

Don't save your R Project to your desktop ; ). Be clear and organised with where and how you store your data and coding files. 

To get '<-', use keyboard shortcut "opt + -" 

To create new chunk in Rmd file, use keyboard shortcut "command + option + I" 

```{r message=FALSE, warning=FALSE}
library(tinytex) #load all necessary packages at the start of the file to track which packages have been loaded to avoid trying to reload them
```

### Introduction to R - Use as a simple calculator 
```{r,results='hide'}
2+1 
6*2
1:30 #Can use R to create vectors 
```

### Section 1.3.8: Variables and Assignments 
```{r,results='hide'}
age<-25 #Use <- to assign a value to an object  
age+1 #uses the value saved as 'age' in future calculations 
#NOTE: Remain consistent with your coding language, if naming things with underscores (e.g., age_in_years), don't start using hyphens etc.
```

#### EXERCISE
```{r,results='hide'}
calc_1<-15+25.1+20.25 #answer is 60.35, assigned to object 'calc_1'
calc_2<-26+51.4+11.07 #answer 88.47, assigned to object 'calc_2' 
calc_1+calc_2 #148.82 
```

### Section 1.3.9: Functions
```{r,results='hide'}
years_old<-25.765
round(years_old) #rounds the value up
floor(years_old) #rounds the value down
```

Each function has arguments, allowing you to be specific about how many decimal places you want to round the number to etc. 
```{r,results='hide'}
round(years_old,2) #25.76, comma after object specifies rounding to 2 decimal places
```

To discover the options for each function, you can query function arguments using args(x), or a question mark for a shortcut to the R help panel 
```{r,results='hide',message=FALSE, warning=FALSE}
?round #takes you to the R help panel for the 'round' function 
args(round) #use in the console (not in the script), will give description of the function 
```

### Section 1.3.10: Debugging
Misconceptions, variables in programs do not work the same way as they do in spreadsheets 

#### EXAMPLE

```{r message=FALSE, warning=FALSE,results='hide'}
grade<-55
total<-grade+10
print(total) #65, as grade was assigned as 55 
grade<-90 
print(total) #value pf total in a spreadsheet will be 100, but in programming a variable holds the value it was assigned (65 in our example). The program still is seeing total as grade (55) + 10 
total<-grade+10 #assigning new value for total 
print(total) #executed in the way it was defined, total = 100 
``` 

Debugging depends on being able to read your code, and is the most effective way of finding errors 

#### EXAMPLE: What will this code do? 
```{r,results='hide'}
p<-2
z<-5
out<-p*z #answer should be 10, given the values assigned previously 
print(out) #print is 10, as expected 
```

#### QUESTION:If the code  out<-p*a was in the same chunk as above, what would happen? 

Error message will appear as 'a' is an undefined object. Would either need to be assigned a numeric value (e.g., a<-8) or corrected to an object that has already been defined (e.g., z or p)

Developing or using testing function is another method, it is a good idea to write short functions that test your results to makre sure they are what they are supposed to be before running your code 

#### EXAMPLE 
```{r,results='hide'}
x<-1 
is.character(x) #output is "FALSE", therefore 'x' is not a character
is.numeric(x) #output is "TRUE", therefore 'x' is numeric 
```

#### EXERCISE: What is wrong with this code? 

#### my_quiz<-c("uno","dos","tres","cuatro","cinco")
#### print(my_quis) 

Quiz is incorrectly spelled "quis", so no object to print 

#### str(my_quiz)

#### len(my_quiz) 

Could not find the function 'len', either incorrectly spelled function (may have meant to type 'length') or need to download/install a certain package to use this function 

### Section 1.3.11: Data Types 

Every object in R has a type, and there are six basic types; character, numeric, integer, logical, complex, and raw. The most common (character, numeric, integer, and logical) are the ones most important to remember. R has built in functions to help examine objects to better understand their type 

class() = what kind of object is it (high-level)?

typeof() = what is the objects data type? 

length() = how long is it? 

attributes() = does it have any metadata?

#### EXERCISE: Try to tick off the top four data types by making a variable and checking what it is 


```{r,results='hide'}
my_numeric_variable<-4.2 
class(my_numeric_variable) #"numeric" 
my_character_variable<-"black"
class(my_character_variable) #"character" 
my_interger_variable<-11L ## don't really understand this one
class(my_interger_variable) #integer 
my_logical_variable<-TRUE
class(my_logical_variable) #logical 
```

### Section 1.3.12: Data Structures 
Elements of the data types above can be combined to form a data structure. Think of it like this - an element is what you would put in a single cell in an excel document, and the whole spreadsheet is the data structure. R has many data structures including atomic vector, list, matrix, data frame, and factors. Vectors are the most common, and it is generally considered the workhorse of R. A vector is a collection of elements that are most commonly of the type character, numeric, integer, logical 

#### EXAMPLES 
```{r,results='hide'}
y<-c(1,2,3) #vector with three numeric elements 
z<-c("Sarah","Tracy","John") #vector with three character elements 
```

##### NOTE: all elements within a vector need to be of the same data type

#### EXERCISE: What happens with you integrate the type of your character vector? 
```{r,results='hide'}
y<-c(1,2,3,numeric) 
z<-c("Sarah","Tracy","John",character) #creates the same vectors, with the data type specified 
```

R also can process lists, and values inside a list can be several different types 
```{r,results='hide'}
x<-list(1,"a",TRUE)
x #will print the list 
#You can retrieve individual elements using double square brackets to reference their index 
x[[2]] #retrieves the second value from the list described above, output shows 'a'
```

### #Section 1.3.12: Data Frames and Tibbles 

R can also handle two-dimensional or 'rectangular' data files, most commonly known as 'data frames'. There are likely the most useful structure for data analysis, with data stored in rows and columns (each columns holds elements of the same type). You can import a data frame from an external file such as .csv or .xlsx file, or you can make a data frame in R by building one (using function 'data-frame'). It lets you make them by adding together some different vectors (lists in this case) and giving each one a column name. 

#### EXAMPLE 

```{r,results='hide'}
my_dataframe<-data.frame(no=c(1,2,3),c("Tracey","John","Pete"),c(TRUE,FALSE,TRUE))
my_dataframe #prints the dataframe
str(my_dataframe) #summary of dataframe with data types, number of variables (based on R's guess - it can be wrong so check!) etc.
#If R has got the data type incorrect, you can change its type using the code below 
my_dataframe$no<-as.factor(my_dataframe$no) #turning data types into factors in column 'no'
str(my_dataframe$no) #shows new data types, with column 'no' being a factor with 3 levels, 1, 2, 3. Factor is a categorical type, so we are telling R that column 1 is storing a category (e.g., a numeric code like a student ID) rather than a continuous number 
```


#### Important Notes 

It is important that each vector is the same length (number of rows) because R won't allow you to have different columns with different lengths. If you have no data to go into a cell, you can use NA. Can use obvious numbers (e.g., 9999) but its better to use NA as there are functions that R can use to determine it (e.g., na.omit or is.na). Try and avoid manually creating dataframes, instead read it in from other files (manually creating greatly increases the risk of transcription errors)

### Section 1.3.14: Packages and Libraries 

Packages are a collection of functions that are made for specific purposes and can be downloaded and used by you. They are generally made by researchers, professors, statisticians, and programmers. To use R packages, you must first download them using install.packages("package_name") and then library(package_name) to load them into your current workspace 

NOTE: once you have installed a package, you won't need to on the same computer again. Best practice is to comment it out (place comment after install.packages command) so that the next time you run it you don't have to install packages again or update them. You can also install multiple packages at once (E.g., install.packages("ggplot","tidyr","dplyr"))

### Section 1.3.15: Coding Best Practices 

Create a section at the top of your script with all the packages needed for your code so you remember which packages you need to load at the start of each coding session. Break your code up into logical, bite-sized sections by using subheadings, new code chunks (if in Markdown) etc. Follow a template - once you have a consistent style for your scripts, write all of your future code using the same style 

#   Workshop Two: Data Visualisation (Thur 21st May, 9-1pm)


```{r Load necessary packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```

### Section 2.16: Load the dataset and examine to see what the dataset looks like
```{r message=FALSE, warning=FALSE, results='hide'}
mpg #built-in dataset in ggplot2, used to teach you the foundations of ggplot package without having to worry about complex datasets etc. 
?mpg #will tell you more about the dataset, including the expanded column names
```

### Section 2.17: Create your first ggplot
ggplot() creates a coordinate system that you can add layers to. The first argument is the dataset, indicated by (data=),then geom_point(), which adds a layer of points to your plot to create a scatterplot (ggplot2 comes with many geom functions to       add different layers to your plot. Each geom function takes a mapping argument, which defines how variables in your dataset are mapped to visual properties. The mapping argument is always paired with aes(), and the x and y arguments within aes() specify which variables to map on the x and y axis 
```{r  }
ggplot(data=mpg)+geom_point(mapping=aes(x=displ,y=hwy))
```

### Section 2.18: Understand the grammar of graphics
#Template : ggplot(data=<DATA>)+<GEOM_FUNCTION>(mapping=aes(<MAPPINGS>))

aes() is the visual property of the objects you've already plotted and inclues things like size, shape, colour etc. 

#### EXAMPLE

```{r message=TRUE, warning=FALSE}
ggplot(data=mpg)+geom_point(mapping=aes(x=displ,y=hwy,colour=class)) #will colour the points based on class 
ggplot(data=mpg)+geom_point(mapping=aes(x=displ,y=hwy,size=class)) #will display class differences as different sized dots 
```

Need to be careful where the '+' is placed when using ggplot, needs to be placed on the top line of code, should not be placed at the start of a new line 

### Section 2.20: Facet and Panel Plots
In ggplot, can use facets to separate plots into subplots/panels 

To facet plots using a single variable, use facet_wrap() (only with discrete variables. facet_wrap() syntax is in the function of a formula, where the ~ dictates which variable you want to subset your data with 

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  facet_wrap(~class,nrow=2) #subsetting data by class, will create a separate panel for each class 
```

If you want to do this for more than variable, use facet_grid(), using two variables with '~' to split them up 
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  facet_grid(drv~cyl) #will facet data by drive type (drv) and number of cylinders (cyl)
```

Use a '.' if you do not want to facet in the rows or column dimension 
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  facet_grid(.~cyl) #ggplot faceted by # of cylinder
```

#### QUESTIONS (answer using ?facet_wrap) 

What does nrow do? number of rows in the plot 

What does ncol do? number of columns in the plot 

What other options control the layout of the individual panels? shrink (shrink scales to fit output stats), as.table (facets laid out like table with highest values in bottom-right) 

### Section 2.21: Fitting simple lines
#To plot points at the coordinate location on the x and y axes, use geom_point() 

To display the same data as a smooth line fit through the points, use geom_smooth()
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ,y=hwy))
```

Can also change the line type and use a variable in the dataset to control it 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  geom_smooth(mapping=aes(x=displ,y=hwy,linetype=drv)) #separates the cars into three lines based on the drive type (drv), and have two different geom functions on the same plot (geom_point() and geom_smooth())
```

Can add the group argument to our gpplot to vary plots by discrete variables 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg)+
  geom_smooth(mapping=aes(x=displ,y=hwy,group=drv,color=drv),
              show.legend=TRUE) #groups by drive type (drv), and are also colored by drive type 
```

#### EXAMPLE: Plotting multiple geoms on a single plot 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  geom_smooth(mapping =aes(x=displ,y=hwy)) #plots both the points and a smooth line on the same graph 
#the code above has the x and y variables separated and indicated in each geom_function, which is not ideal if you are wanting to change a variable as you will have to change it in multiple locations - to avoid this, you can specify the x and y variable in the ggplot() argument, which will apply these variables to each geom_function in the plot 
ggplot(data=mpg,mapping=aes(x=displ,y=hwy))+
  geom_point()+
  geom_smooth() #generates the same plot as above, but the variables are defined in the ggplot() argument, not in the geom_function()
```

You can use mapping in specific layers to enable the display the different aesthetics in different layers 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy))+
  geom_point(mapping=aes(color=class))+
  geom_smooth() #differentiates class by coloured points 
```

Can also be used if you want to specify different data for each layer - example below uses a filter (class="subcompact") to select a subset of the data and plot only that subset 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy))+
  geom_point(mapping=aes(color=class))+
  geom_smooth(data=filter(mpg,class=="subcompact",se=TRUE)) #plot shows points by colour, and the smooth line has been filtered by cars in the "subcompact" class 
```

#### EXERCISE


##### QUESTION ONE: 

What geom would you use to draw a ... 

  Line chart? geom_line()
  
  Boxplot? geom_boxplot()
  
  Histogram? geom_histogram()
  
  Area chart? geom_area()
  
##### QUESTION TWO: 
Run this code in your head first and visualise what the output will look like. Will the two graphs look different? Why or why not? 

PREDICTION: These two graphs will look the same as they are both plotting the same things with the same geom_functions. The only difference is the first code has specified the x and y variables in ggplot argument, whereas the second code has specificed x and y for each geom_function. 
```{r message=FALSE, warning=FALSE}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy))+
  geom_point()+
  geom_smooth() #variables are specified in ggplot() argument
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))+
  geom_smooth(mapping =aes(x=displ,y=hwy)) #variables are specified for each geom_function()
```

### Section 2.22.1: Plotting Statistics
```{r }
#using the 'diamonds' dataset embedded in ggplot2 
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut)) #shows more diamonds are available with high quality cuts than low quality cuts, geom_bar has placed data into 'bins' and then plotted the counts within each bin on the x-axis (same as histograms and frequency polygons)
    #smoothers fit a model to your data and then plot predictions from the model 
    #boxplots compute a robust summary of the distribution and then display a specially formatted box 
```

The algorithm used to calculate new values for a graph is called a stat (short for statistical transformation)

You can generally use geoms and stats interchangeably 

#### EXAMPLE
```{r message=FALSE, warning=FALSE}
ggplot(data=diamonds)+
  stat_count(mapping=aes(x=cut)) #creates same plot as using geom_bar, but uses stat_count instead 
```


### Section 2.22.2: Overriding Defaults in ggplot2 
You may want to override a default stat, such as from a count/summary to identity (which is the raw value of a variable)
```{r }

demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
ggplot(data=demo)+
  geom_bar(mapping=aes(x=cut,y=freq),stat="identity")
```

You can also override a default mapping from transformed variables to aesthetics. 

#### EXAMPLE BELOW

You could display a bar chart of the proportion of your total diamond dataset rather than count 
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut,y=after_stat(prop),group=1)) #x-axis is a proportion of total data set, not a count 
```

### Section 2.22.3: Plotting statistical details 
It is good practice to be transparent about transformation in your plot and any uncertainties or limitations of your data, can do this using stat_summary()
```{r Section 2.22.3}
ggplot(data=diamonds)+
  stat_summary(
    mapping=aes(x=cut,y=depth),
    fun.min=min, 
    fun.max=max,
    fun=median
  )
```

### Section 2.23: Aesthetic Adjustments 
Can use aesthetics to colour or fill to change aspects of the bar colours
```{r Section 2.23: Aesthetic Adjustments }
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut,colour=cut)) #outlines the different bars (binned by cut) in different colors 
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut,fill=cut)) #fills the different bars (binned by cut)
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut,fill=clarity)) #colored based on clarity (another variable in the data set), done behind the scenes using a position argument 

```

Position adjustments allow you to customise your plot in three ways; identity (raw data), fill (change heights), and dodge (forces ggplot2 to not put things on top of each other)

If using position = "identity", you can place each object exactly where it falls in the context of the graph. This is vital for points charts (e.g., scatter plots), but messy for others (e.g., bar plots). 

Can alter the bar aesthetic
```{r}
ggplot(data=diamonds,mapping=aes(x=cut,fill=clarity))+
  geom_bar(alpha=1/5,position="identity") #the alpha = 1/5 changes the transparency (1/5 = 20% opacity)
ggplot(data=diamonds,mapping=aes(x=cut,colour=clarity))+
  geom_bar(fill=NA,position="identity") #outlines the bars without filling with color 
```

Position = "fill" works like stacking, but makes each set of stacked bars the same height
```{r}
ggplot(data=diamonds,mapping=aes(x=cut,fill=clarity))+
  geom_bar(position="fill")
```


Position = "dodge" places overlapping objects directly beside one another 
```{r}
ggplot(data=diamonds)+
  geom_bar(mapping=aes(x=cut,fill=clarity),position="dodge") #each cut type now has a separate bar for each clarity type (not stacked on top of each other anymore)
```

Position = "jitter" slightly moves points (by adding a small amount of random. noise to each point) to avoid overplotting when points overlap. This is useful for scatterplots, but not barplots 
```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy),position="jitter") 
```

### Section 2.24: Layered Grammar of Graphics 

Updated template for ggplot below: 
ggplot(data=DATA) + GEOM_FUNCTION(mapping = aes(MAPPINGS),stat = STAT, position = POSITION) +FACET_FUNCTION


