---
title: "MB5370 Module 1 Workshop Three"
author: "Caitlin Buckle"
date: "2026-05-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load necessary packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
```

### Section 
```{r message=FALSE, warning=FALSE}

```

### How to Source Code from R Script in Rmd (without having to copy and paste)
To source the code in our workshop script we can use the source() function in R. This will execute the code in the specified file and make any functions or variables defined in that file available in our current R session. 
#### EXAMPLE CODE: source ("../code/Module_1.R")
the "../" will instruct R to go 'up' one folder (or out of one folder) and into the next required folder "/code/" in this example. 
#### NOTE: the source() function will not work unless the code in your script has no issues or errors. If the R-script file is complete, the source should run. 
```{r message=FALSE, warning=FALSE}
source("Module_1.R") #The R-script file is in the same folder on my computer as the R project file, therefore I do not need to specify having to exit a folder and into another one (therefore don't need to include "../" in my code). 
print(plot) # in R-script, the ggplot will appear in the plot window automatically, but if you don't have it specified for R to print the plot, you will not see anything when you source the plot from your R-script in Rmd. 
```

### Section 3.4: Setting Up Git and Github 
Git was downloaded via Mac Ports (using older version as laptop is not updated yet)
Used command " sudo port install git " into Mac terminal to install 
Created a token, selected no expiry so won't have to repeat these steps during the course 
Followed the instructions in the workshop manual to install and get Git communicating with Github 
Created a new project with the class repo for Module_1 

#### NOTE: important to consider how you structure your files in finder, can follow the structure outlined in the manual as an example 
#### EXAMPLE :
MB5370-Programming/ <- main folder with everything inside 
│
├── MB5370-Programming.Rproj
├── data/       <- For raw, unedited data files (CSVs, shapes, etc.)
├── code/       <- For dirty/heavy data manipulation code
├── output/     <- For processed tables, figures, and saved objects
└── docs/       <- For your rendered write-ups and final documents

### Section 3.6: Track Your Own Project 
Three main options for tracking your own project and pushing it to Github (all within RStudio and all within the specific project)
#### Option One: call 'usethis::use_git()' in the R console 
#### Option Two: 
Using tools. In RStudio, go to Tools > Project Options... > Git/SVN. Under 'version control system', select 'Git'. Confirm New Git Repository > Yes 
#### Options Three: Using the terminal 
Click on the terminal pane at the bottom. With working directory set to the projects directory, run git init. 

Used 'usethis' option 

#### NOTE: On Macs, it you cannot see the .gitignore folders, do command + shift + . and it will show all hidden/private folders in Finder 



